Constructing Quotient Inductive-Inductive Types

### 2018-11-09 paper.tex
Add ACM info
Remove redundant ACMsetup, add submission response page
Split appendix from paper

\usepackage{proof}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{pdflscape}
\usepackage{multirow,bigdelim}
\input{abbrevs.tex}
\usepackage{tikz}
\usetikzlibrary{arrows,
                calc,
                decorations.pathreplacing,
                calligraphy,% had to be after decorations.pathreplacing
                matrix,
                positioning
                }

@@

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}

@@

@@

In the following table we summarize the operations which we define in
Sections \ref{sec:algebras}--\ref{sec:eliminator}. $\Gamma$ denotes a
signature. The full definitions of these operations are given in
Appendix \ref{sec:app}.

@@

Our metatheory is Martin-L{\"o}f type theory with functional
extensionality and uniqueness of identity proofs (UIP). In this
section we describe the notation used in this paper and the accompanying
Agda formalisation.

@@

The Agda formalisation has been checked using Agda 2.5.4. It covers
sections \ref{sec:syntax}, \ref{sec:syntaxfunctions},
\ref{sec:algebras}, the definition of homomorphisms from
\ref{sec:recursor}, and also \ref{sec:eliminator},
\ref{sec:definingcwf}, \ref{sec:cwfalg}, \ref{sec:initind}, and part
of \ref{sec:cwfmodel}. The formalisation also includes additional
documentation concerning technicalities.

@@

In the following sections, we will define functions by induction on
the syntax of the theory of signatures. This amounts to saying what
$\Con$, $\Ty$, $\Tms$ and $\Tm$ are mapped to and providing cases for
all of the 35 constructors from $\cdot$ to ${\appm[]}$. Examples of
full definitions of some of these functions are given in Appendix
\ref{sec:app}, the simplest one being the standard interpretation
$\blank^\A$.

@@

The above four lines are the specification of the operation
$\blank^\A$. Its definition amounts to describing what it does on all
the constructors of the theory of signatures. We start by saying that
$\blank^\A$ on the empty context returns the unit type $\cdot^\A
:\equiv \top$, context extension is mapped to $\Sigma$, that is
$(\Gamma\rhd A)^\A:\equiv (\gamma:\Gamma^\A)\times
A^\A\,\gamma$. $\pi_1$ and $\pi_2$ are interpreted as first and second
projections, respectively, so a variable projects out the
corresponding component, e.g. $2^\A\,(\gamma,\alpha, \alpha',\alpha'')
\equiv \alpha$. $\U$ is interpreted as $\Set$, that is, $\U^\A\,\gamma
:\equiv \Set$, a type coming from a code uses the interpretation of
the code: $(\El\,a)^\A\,\gamma :\equiv a^\A\,\gamma$ which has the
right type as $\U^\A\,\gamma \equiv \Set$. Both function spaces are
mapped to metatheoretic function space: $(\Pi\,a\,B)^\A\,\gamma
:\equiv (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha)$ and
$(\Pim\,T\,B)^\A\,\gamma :\equiv
(\alpha:T)\ra(B\,\alpha)^\A\,\gamma$. The identity type is mapped to
the metatheoretic identity type: $(\Id\,a\,t\,u)^\A\,\gamma :\equiv
(t^\A\,\gamma=u^\A\,\gamma)$. The standard interpretation justifies
equality reflection when the metatheory supports functional
extensionality: we need to provide $(\reflect\,e)^\A : t^\A = u^\A$,
and we have $e^\A:(\gamma:\Gamma^\A)\ra
t^\A\,\gamma=u^\A\,\gamma$. Hence $(\reflect\,e)^\A$ is just given by
$\funext$. All the other equality constructors are interpreted by
$\refl$. The full definition for every constructor of the theory of
signatures is given in Appendix \ref{sec:app}.

@@

We refer the interested reader to Appendix \ref{sec:app} for the full
definition of this and later operations.

@@

Displayed algebras are given by the unary logical predicate
interpretation \cite{bernardy12parametricity,atkey}. Contexts become
predicates over algebras, types become dependent predicates, and
substitutions and terms produce witnesses of the logical predicates
once the logical predicate is witnessed at the context.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\D && : \Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\D && : \Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\D && : \Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma) \\
  & (t:\Tm\,\Gamma\,A)^\D && : (\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)
\end{alignat*}
Here the interpretation of $\U$, $\El$ and $\Pi$ are the usual ones,
$\U$ becomes predicate space ($\U^\D\,\gamma^D\,T :\equiv T\ra\Set$),
$\El\,a$ is just a witness of the predicate for $a$, and function
space is interpreted as the predicate which expresses preservation of
logical predicates. The details can be found in Appendix
\ref{sec:app}.

@@

For the eliminator we fix a signature $\Omega$ and a displayed algebra
over the initial algebra $\omega^D:\Omega^\D\,\con_\Omega$. Then we
define the operation $\blank^\E$ by induction on the syntax. The
specification is as follows.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Gamma^\S\,(\nu^\A\,\con)\,(\nu^\D\,\omega^D) \\
  & (A:\Ty\,\Gamma)^\E && : (\nu:\Tms\,\Omega\,\Gamma)(t:\Tm\,\Omega\,(A[\nu]))\ra A^\S\,(\Gamma^\E\,\nu)\,(t^\A\,\con)\,(t^\D\,\omega^D) \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Delta^\E\,(\sigma\circ\nu) = \sigma^\S\,(\Gamma^\E\,\nu) \\
  & (t:\Tm\,\Gamma\,A)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra A^\E\,\nu\,(t[\nu]) = t^\S\,(\Gamma^\E\,\nu)
\end{alignat*}
The definition is analogous to that of $\blank^\R$. One difference is
that for $\U$ we have an additional transport. The type that we need
is $(\alpha:a^\A\,\con)\ra (a^\D\,\omega^D\,\alpha)$. The expression
$\coe\,({a^\C\,\id}^{-1})\,\alpha$ has type
$\Tm\,\Omega\,(\El\,a)$. Now, instead of applying the standard
interpretation, we apply the logical predicate interpretation:
$(\coe\,({a^\C\,\id}^{-1})\,\alpha)^\D\,\omega^D$ has type
$a^\D\,\omega^\D\,(\alpha^\A\,\con)$ which is almost right, so we need
to coerce along $\alpha^\C\,\id$ which witnesses $\alpha^\A\,\con =
\alpha$, and we are finished. For details see Appendix \ref{sec:app}
and the Agda formalisation.

@@

\draw (m-2-2.north west) rectangle (m-17-6.south west) node[pos=.5] {\rotatebox{90}{Appendix \ref{sec:app}}};

\draw (m-2-15.north west) rectangle (m-17-18.south west) node[pos=.5] {\rotatebox{90}{Appendix \ref{sec:app}}};

@@

\draw (m-2-2.north west) rectangle (m-5-14.south west) node[pos=.5] {Appendix \ref{sec:motives}};

\draw (m-9-2.north west) rectangle (m-10-14.south west) node[pos=.5] {Appendix \ref{sec:motives}};
\draw (m-12-2.north west) rectangle (m-12-14.south west) node[pos=.5] {Appendix \ref{sec:motives}};

@@

There are two ways to present the model: by rows or by columns of the
table. Describing the model by rows is the usual way of saying how
contexts are given in the model, then how types are given, then
substitutions, terms, the empty context, context extension and so on.
Describing the model by columns means defining operations which
interpret the syntax, and later operations can depend on previous
ones. In Sections \ref{sec:algebras}--\ref{sec:eliminator} we used the
column-based method to describe the first four columns of this
table. These were given by $\blank^\A$, $\blank^\D$, $\blank^\M$, and
$\blank^\S$ (their full definition can be found in Appendix
\ref{sec:app}).

In this section we follow the row-based approach and describe the rows
of the model informally, while giving the full definition of the first
4 rows ($\Con$, $\Ty$, $\Tms$, $\Tm$) and the rows for $\U$, $\El$ and
$\Pi$ in Appendix \ref{sec:motives}. These are the most interesting
parts of the construction. The rest of the construction is tedious and
not very enlightening. The first 4 columns are fully formalised in
Agda and most parts of the category-columns from the CwF-columns are
also formalised.

@@

%% Appendix
\appendix
\begin{landscape}
\section{Definitions of the operations from Sections \ref{sec:algebras}--\ref{sec:eliminator}}
\label{sec:app}

In this appendix we give the full definitions of the operations
$\blank^\A$, $\blank^\C$ (Section \ref{sec:algebras}), $\blank^\M$,
$\blank^\R$ (Section \ref{sec:recursor}), $\blank^\D$, $\blank^\S$,
$\blank^\E$ (Section \ref{sec:eliminator}).

{\footnotesize
\begin{alignat*}{10}
  & \text{\textbf{Syntax}}                                                                             && \rlap{$\text{\textbf{Algebras}}$} &&                                                      && \rlap{$\text{Assuming $\Omega:\Con$, the \textbf{initial $\Omega$-algebra} is given by $\con_\Omega:\equiv \Omega^\C\,\id$}$}                                                                                          \\
  & \Gamma:\Con                                                                                        && \Gamma^\A && : \Set                                                                             && \Gamma^\C && : \Tms\,\Omega\,\Gamma\ra\Gamma^\A                                                                                          \\
  & A:\Ty\,\Gamma                                                                                      && A^\A && : \Gamma^\A\ra\Set                                                                      && A^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra\Tm\,\Omega\,(A[\nu])\ra A^\A\,(\Gamma^\C\,\nu)                                                   \\
  & \sigma:\Tms\,\Gamma\,\Delta                                                                        && \sigma^\A && : \Gamma^\A\ra\Delta^\A                                                            && \sigma^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Delta^\C\,(\sigma\circ\nu)=\sigma^\A\,(\Gamma^\C\,\nu)                                     \\
  & t:\Tm\,\Gamma\,A                                                                                   && t^\A && : (\gamma:\Gamma^\A)\ra A^\A\,\gamma                                                    && t^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra A^\C\,\nu\,(t[\nu])=t^\A\,(\Gamma^\C\,\nu)                                                       \\
  & \cdot:\Con                                                                                         && \cdot^\A && :\equiv \top                                                                        && \cdot^\C\,\nu && :\equiv \tt                                                                                                             \\
  & \Gamma\rhd A:\Con                                                                                  && (\Gamma\rhd A)^\A && :\equiv (\gamma:\Gamma^\A)\times A^\A\,\gamma                              && (\Gamma\rhd A)^\C\,\nu && :\equiv (\Gamma^\C\,(\pi_1\,\nu),A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu))                                             \\
  & (A:\Ty\,\Delta)[\sigma:\Tms\,\Gamma\,\Delta]:\Ty\,\Gamma                                           && (A[\sigma])^\A\,\gamma && :\equiv A^\A\,(\sigma^\A\,\gamma)                                     && (A[\sigma])^\C\,\nu\,t && :\equiv \tr_{A^\A}\,(\sigma^\C\,\nu)\,(A^\C\,(\sigma\circ\nu)\,t)                                              \\
  & \id : \Tms\,\Gamma\,\Gamma                                                                         && \id^\A\,\gamma && :\equiv \gamma                                                                && \id^\C\,\nu && : \Gamma^\C\,\nu = \Gamma^\C\,\nu                                                                                                             \\
  & (\sigma:\Tms\,\Theta\,\Delta)\circ(\delta:\Tms\,\Gamma\,\Theta):\Tms\,\Gamma\,\Delta               && (\sigma\circ\delta)^\A\,\gamma && :\equiv \sigma^\A\,(\delta^\A\,\gamma)                        && (\sigma\circ\delta)^\C\,\nu && : \Delta^\C\,(\sigma\circ\delta\circ\nu) \overset{\sigma^\C\,(\delta\circ\nu)}{=} \sigma^\A\,(\Theta^\C\,(\delta\circ\nu)) \overset{\delta^\C\,\nu}{=} \sigma^\A\,(\delta^\A\,(\Gamma^\C\,\nu)) \\
  & \epsilon : \Tms\,\Gamma\,\cdot                                                                     && \epsilon^\A\,\gamma && :\equiv \tt                                                              && \epsilon^\C\,\nu && : \tt = \tt                                                                                                        \\
  & (\sigma:\Tms\,\Gamma\,\Delta) , (t:\Tm\,\Gamma\,(A[\sigma])):\Tms\,\Gamma\,(\Delta\rhd A)\hspace{1em}          && (\sigma, t)^\A\,\gamma && :\equiv (\sigma^\A\,\gamma, t^\A\,\gamma)                             && (\sigma,t)^\C\,\nu && : (\Gamma^\C\,(\sigma\circ\nu), A^\C\,(\sigma\circ\nu)\,(t[\nu])) \overset{\sigma^\C\,\nu, t^\C\,\nu}{=} (\sigma^\A\,(\Gamma^\C\,\nu), t^\A\,(\Gamma^\C\,\nu))                                                                        \\
  & \pi_1\,(\sigma:\Tms\,\Gamma\,(\Delta\rhd A)):\Tms\,\Gamma\,\Delta                                  && (\pi_1\,\sigma)^\A\,\gamma && :\equiv \proj_1\,(\sigma^\A\,\gamma)                              && (\pi_1\,\sigma)^\C\,\nu && : \Delta^\C\,(\pi_1\,(\sigma\circ\nu)) \overset{\sigma^\C\,\nu}{=} \proj_1\,(\sigma^\A\,(\Gamma^\C\,\nu))                                                                        \\
  & \pi_2\,(\sigma:\Tms\,\Gamma\,(\Delta\rhd A)):\Tm\,\Gamma\,(A[\pi_1\,\sigma])                       && (\pi_2\,\sigma)^\A\,\gamma && :\equiv \proj_2\,(\sigma^\A\,\gamma)                              && (\pi_2\,\sigma)^\C\,\nu && : A^\C\,(\pi_1\,(\sigma\circ\nu))\,(\pi_2\,(\sigma\circ\nu)) \overset{\sigma^\C\,\nu}{=} \proj_2\,(\sigma^\A\,(\Gamma^\C\,\nu))                                                                           \\
  & (t:\Tm\,\Delta\,A)[\sigma:\Tms\,\Gamma\,\Delta]:\Tm\,\Gamma\,(A[\sigma])                           && (t[\sigma])^\A\,\gamma && :\equiv t^\A\,(\sigma^\A\,\gamma)                                     && (t[\sigma])^\C\,\nu && : A^\C\,(\sigma\circ\nu)\,(t [\sigma\circ\nu]) \overset{t^\C(\sigma\circ\nu)}{=} t^\A\,(\delta^\C\,(\sigma\circ\nu))\overset{\sigma^\C\,\nu}{=}t^\A\,(\sigma^\A(\Gamma^\C\,\nu))  \\
  & [\id] : A[\id] = A                                                                                 && [\id]^\A && :\equiv \refl                                                                       && [\id]^\C && : A^\C\,\nu\,t = A^\C\,\nu\,t                                                                                                                \\
  & [\circ] : A[\sigma \circ \delta] = A[\sigma][\delta]                                               && [\circ]^\A && :\equiv \refl                                                                     && [\circ]^\C && : A^\C\,(\sigma\circ\delta\circ\nu)\,t = A^\C\,(\sigma\circ\delta\circ\nu)\,t                                                       \\
  & \ass : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu)                           && \ass^\A && :\equiv \refl                                                                        && \ass^\C && :\equiv \UIP                                                                                              \\
  & \idl : \id\circ\sigma = \sigma                                                                     && \idl^\A && :\equiv \refl                                                                        && \idl^\C && :\equiv \UIP                                                                                              \\
  & \idr : \sigma\circ\id = \sigma                                                                     && \idr^\A && :\equiv \refl                                                                        && \idr^\C && :\equiv \UIP                                                                                              \\
  & {\cdot\eta} : \{\sigma : \Tms\,\Gamma\,\cdot\} \ra \sigma = \epsilon                               && {\cdot\eta}^\A && :\equiv \refl                                                                 && {\cdot\eta}^\C && :\equiv \UIP                                                                                       \\
  & {\rhd\beta_1} : \pi_1\,(\sigma, t) = \sigma                                                        && {\rhd\beta_1}^\A && :\equiv \refl                                                               && {\rhd\beta_1}^\C&& :\equiv \UIP                                                                                      \\
  & {\rhd\beta_2} : \pi_2\,(\sigma, t) = t                                                             && {\rhd\beta_2}^\A && :\equiv \refl                                                               && {\rhd\beta_2}^\C && :\equiv \UIP                                                                                     \\
  & {\rhd\eta} : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma                                               && {\rhd\eta}^\A && :\equiv \refl                                                                  && {\rhd\eta}^\C&& :\equiv \UIP                                                                                         \\
  & {,\circ} : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta])                                 && {,\circ}^\A && :\equiv \refl                                                                    && {,\circ}^\C && :\equiv \UIP                                                                                          \\
  & \U : \Ty\,\Gamma                                                                                   && \U^\A\,\gamma && :\equiv \Set                                                                   && \U^\C\,\nu\,a && :\equiv \Tm\,\Omega\,(\El\,a)                                                                                        \\
  & \El\,(a: \Tm\,\Gamma\,\U): \Ty\,\Gamma                                                             && (\El\,a)^\A\,\gamma && :\equiv a^\A\,\gamma                                                     && (\El\,a)^\C\,\nu\,t    && :\equiv \coe\,(a^\C\,\nu : \Tm\,\Omega\,(\El\,a)=a^\A\,(\Gamma^\C\,\nu))\,t                                                                                   \\
  & {\U[]} : \U[\sigma] = \U                                                                           && {\U[]}^\A && :\equiv \refl                                                                      && {\U[]}^\C && : \Tm\,\Omega\,a=\Tm\,\Omega\,a                                                  \\
  & {\El[]} : (\El\,a)[\sigma] = \El\,(a[\sigma])                                                      && {\El[]}^\A && :\equiv \refl                                                                     && {\El[]}^\C && : t = t                                                                                        \\
  & \Pi\,(a:\Tm\,\Gamma\,\U)(B:\Ty\,(\Gamma\rhd\El\,a)):\Ty\,\Gamma                                    && (\Pi\,a\,B)^\A\,\gamma && :\equiv (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma, \alpha) \hspace{1em}  && (\Pi\,a\,B)^\C\,\nu\,t && :\equiv \lambda\alpha.B^\C\,(\nu,\coe\,({a^\C\,\nu^{-1}})\,\alpha)\,(t\app \coe\,({a^\C\,\nu^{-1}})\,\alpha)   \\
  & \APP\,(t:\Tm\,\Gamma\,(\Pi\,a\,B)):\Tm\,(\Gamma\rhd\El\,a)\,B\hspace{1em}                          && (\APP\,t)^\A\,(\gamma,\alpha) && :\equiv t^\A\,\gamma\,\alpha                                    && (\APP\,t)^\C\,\nu    && : B^\C\,\nu\,((\APP\,t)[\nu]) \overset{t^\C\,(\pi_1\,\nu)}{=} t^\A\,(\Gamma^\C\,(\pi_1\,\nu))\,(\pi_2\,\nu)                                                                                                          \\
  & {\Pi[]} : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow])                              && {\Pi[]}^\A && :\equiv \refl                                                                    &&  {\Pi[]}^\C && : \lambda\alpha.B^\C\,(\sigma\circ\nu,\alpha)\,(t\app\alpha) = \lambda\alpha.B^\C\,(\sigma\circ\nu,\alpha)\,(t\app\alpha)                                                                                                                   \\
  & {\APP[]} : (\APP\,t)[\sigma\uparrow] = \APP\,(t[\sigma])                                           && {\APP[]}^\A && :\equiv \refl                                                                    && {\APP[]}^\C && :\equiv \UIP                                                                                          \\
  & \Id\,(a:\Tm\,\Gamma\,\U)\,(t\,u:\Tm\,\Gamma\,(\El\,a)):\Ty\,\Gamma                                 && (\Id\,a\,t\,u)^\A\,\gamma && :\equiv (t^\A\,\gamma=u^\A\,\gamma)                                && (\Id\,a\,t\,u)^\C\,\nu\,e && : t^\A\,(\Gamma^\C\,\nu)\overset{t^\C\,\nu}{=} t[\nu] \overset{\reflect\,e}{=} u[\nu] \overset{u^\C\,\nu}{=} u^\A\,(\Gamma^\C\,\nu)                                                                                                    \\
  & \reflect\,(e:\Tm\,\Gamma\,(\Id\,a\,t\,u)): t = u                                                   && (\reflect\,e)^\A && :\equiv \funext\,e^\A                                                       && (\reflect\,e)^\C && :\equiv \UIP                                                                                                              \\
  & {\Id[]} : (\Id\,a\,t\,u)[\sigma] = \Id\,(a[\sigma])\,(t[\sigma])\,(u[\sigma])                      && {\Id[]}^\A && :\equiv \refl                                                                     && {\Id[]}^\C && :\equiv \UIP                                                                                                                    \\
  & \Pim\,(T:\Set)\,(B:T\ra\Ty\,\Gamma):\Ty\,\Gamma                                                    && (\Pim\,T\,B)^\A\,\gamma && :\equiv (\alpha:T)\ra (B\,\alpha)^\A\,\gamma                         && (\Pim\,T\,B)^\C\,\nu\,t && :\equiv \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha)                                                                                                      \\
  & (t:\Tm\,\Gamma\,(\Pim\,T\,B))\appm(\alpha:T):\Tm\,\Gamma\,(B\,\alpha)                              && (t\appm\alpha)^\A\,\gamma && :\equiv t^\A\,\gamma\,\alpha                                       && (t\appm\alpha)^\C\,\nu    && : (B\,\alpha)^\C\,\nu\,(t[\nu]\appm\alpha)\overset{t^\C\,\nu}{=}t^\A\,(\Gamma^\C\,\nu)\,\alpha                                                                                                      \\
  & {\Pim[]} : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma])                    && {\Pim[]}^\A && :\equiv \refl                                                                    && {\Pim[]}^\C && : \lambda\alpha.(B\,\alpha)^\C\,(\sigma\circ\nu)\,(t\appm\alpha) =                                                                                                                    \\
  & && && && && \hspace{1.5em} \lambda\alpha.(B\,\alpha)^\C\,(\sigma\circ\nu)\,(t\appm\alpha)(\Pim\,T\,(\lambda\alpha.(B\,\alpha)[\sigma]))^\C\,\nu\,t                                                                                                                   \\
  & {\appm[]} : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha                               && {\appm[]}^\A && :\equiv \refl                                                                   && {\appm[]}^\C && :\equiv \UIP
\end{alignat*}

\newpage

\begin{alignat*}{10}
  & \text{\textbf{Syntax}}                                                                                && \rlap{$\text{\textbf{Homomorphisms}}$} &&                                                                                                                                       \\
  & \Gamma:\Con                                                                                           && \Gamma^\M && : \Gamma^\A\ra\Gamma^\A\ra\Set                                                                                                                                     \\
  & A:\Ty\,\Gamma                                                                                         && A^\M && : \Gamma^\M\,\gamma^\0\,\gamma^\1\ra A^\A\,\gamma^\0\ra A^\A\,\gamma^\1\Set                                                                                             \\
  & \sigma:\Tms\,\Gamma\,\Delta                                                                           && \sigma^\M && : \Gamma^\M\,\gamma^\0\,\gamma^\1\ra\Delta^\M\,(\sigma^\A\,\gamma^\0)\,(\sigma^\A\,\gamma^\1)                                                                      \\
  & t:\Tm\,\Gamma\,A                                                                                      && t^\M && : (\gamma^M:\Gamma^\M\,\gamma^\0\,\gamma^\1)\ra A^\M\,\gamma^M\,(t^\A\,\gamma^\0)\,(t^\A\,\gamma^\1)                                                                     \\
  & \cdot:\Con                                                                                            && \cdot^\M\,\tt\,\tt && :\equiv \top                                                                                                                                              \\
  & \Gamma\rhd A:\Con                                                                                     && (\Gamma\rhd A)^\M\,(\gamma^\0,\alpha^\0)\,(\gamma^\1,\alpha^\1) && :\equiv (\gamma^M:\Gamma^\M\,\gamma^\0\,\gamma^\1)\times A^\M\,\gamma^M\,\alpha^\0\,\alpha^\1                \\
  & (A:\Ty\,\Delta)[\sigma:\Tms\,\Gamma\,\Delta]:\Ty\,\Gamma                                              && (A[\sigma])^\M\,\gamma^M\,\alpha^\0\,\alpha^\1 && :\equiv A^\M\,(\sigma^\M\,\gamma^M)\,\alpha^\0\,\alpha^\1                                                                     \\
  & \id : \Tms\,\Gamma\,\Gamma                                                                            && \id^\M\,\gamma^M && :\equiv \gamma^m                                                                                                                                            \\

>>>>

\usepackage{proof}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{pdflscape}
\usepackage{multirow,bigdelim}
\usepackage{hyperref}
\input{abbrevs.tex}
\usepackage{tikz}
\usetikzlibrary{arrows,
                calc,
                decorations.pathreplacing,
                calligraphy,% had to be after decorations.pathreplacing
                matrix,
                positioning
                }

@@

@@

%%% The following is specific to POPL '19 and the paper
%%% 'Constructing Quotient Inductive-Inductive Types'
%%% by Ambrus Kaposi, András Kovács, and Thorsten Altenkirch.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3290315}
\acmYear{2019}
\copyrightyear{2019}
\acmJournal{PACMPL}
\acmVolume{3}
\acmNumber{POPL}
\acmArticle{2}
\acmMonth{1}

@@

In the following table we summarize the operations which we define in
Sections \ref{sec:algebras}--\ref{sec:eliminator}. $\Gamma$ denotes a
signature. The full definitions of these operations are given in
Appendix A.

@@

Our metatheory is Martin-L{\"o}f type theory with functional
extensionality and uniqueness of identity proofs (UIP). In this
section we describe the notation used in this paper and the
accompanying technical appendix and Agda formalisation.

@@

The Agda formalisation has been checked using Agda 2.5.4. It is
available at
\url{https://bitbucket.org/akaposi/finitaryqiit/src/master/}, together
with a technical appendix. The Agda formalisation covers sections
\ref{sec:syntax}, \ref{sec:syntaxfunctions}, \ref{sec:algebras}, the
definition of homomorphisms from \ref{sec:recursor}, and also
\ref{sec:eliminator}, \ref{sec:definingcwf}, \ref{sec:cwfalg},
\ref{sec:initind}, and part of \ref{sec:cwfmodel}. The formalisation
also includes additional documentation concerning technicalities.

@@

In the following sections, we will define functions by induction on
the syntax of the theory of signatures. This amounts to saying what
$\Con$, $\Ty$, $\Tms$ and $\Tm$ are mapped to and providing cases for
all of the 35 constructors from $\cdot$ to ${\appm[]}$. Examples of
full definitions of some of these functions are given in Appendix A,
the simplest one being the standard interpretation $\blank^\A$.

@@

The above four lines are the specification of the operation
$\blank^\A$. Its definition amounts to describing what it does on all
the constructors of the theory of signatures. We start by saying that
$\blank^\A$ on the empty context returns the unit type $\cdot^\A
:\equiv \top$, context extension is mapped to $\Sigma$, that is
$(\Gamma\rhd A)^\A:\equiv (\gamma:\Gamma^\A)\times
A^\A\,\gamma$. $\pi_1$ and $\pi_2$ are interpreted as first and second
projections, respectively, so a variable projects out the
corresponding component, e.g. $2^\A\,(\gamma,\alpha, \alpha',\alpha'')
\equiv \alpha$. $\U$ is interpreted as $\Set$, that is, $\U^\A\,\gamma
:\equiv \Set$, a type coming from a code uses the interpretation of
the code: $(\El\,a)^\A\,\gamma :\equiv a^\A\,\gamma$ which has the
right type as $\U^\A\,\gamma \equiv \Set$. Both function spaces are
mapped to metatheoretic function space: $(\Pi\,a\,B)^\A\,\gamma
:\equiv (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha)$ and
$(\Pim\,T\,B)^\A\,\gamma :\equiv
(\alpha:T)\ra(B\,\alpha)^\A\,\gamma$. The identity type is mapped to
the metatheoretic identity type: $(\Id\,a\,t\,u)^\A\,\gamma :\equiv
(t^\A\,\gamma=u^\A\,\gamma)$. The standard interpretation justifies
equality reflection when the metatheory supports functional
extensionality: we need to provide $(\reflect\,e)^\A : t^\A = u^\A$,
and we have $e^\A:(\gamma:\Gamma^\A)\ra
t^\A\,\gamma=u^\A\,\gamma$. Hence $(\reflect\,e)^\A$ is just given by
$\funext$. All the other equality constructors are interpreted by
$\refl$. The full definition for every constructor of the theory of
signatures is given in Appendix A.

@@

We refer the interested reader to Appendix A for the full
definition of this and later operations.

@@

Displayed algebras are given by the unary logical predicate
interpretation \cite{bernardy12parametricity,atkey}. Contexts become
predicates over algebras, types become dependent predicates, and
substitutions and terms produce witnesses of the logical predicates
once the logical predicate is witnessed at the context.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\D && : \Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\D && : \Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\D && : \Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma) \\
  & (t:\Tm\,\Gamma\,A)^\D && : (\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)
\end{alignat*}
Here the interpretation of $\U$, $\El$ and $\Pi$ are the usual ones,
$\U$ becomes predicate space ($\U^\D\,\gamma^D\,T :\equiv T\ra\Set$),
$\El\,a$ is just a witness of the predicate for $a$, and function
space is interpreted as the predicate which expresses preservation of
logical predicates. The details can be found in Appendix A.

@@

For the eliminator we fix a signature $\Omega$ and a displayed algebra
over the initial algebra $\omega^D:\Omega^\D\,\con_\Omega$. Then we
define the operation $\blank^\E$ by induction on the syntax. The
specification is as follows.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Gamma^\S\,(\nu^\A\,\con)\,(\nu^\D\,\omega^D) \\
  & (A:\Ty\,\Gamma)^\E && : (\nu:\Tms\,\Omega\,\Gamma)(t:\Tm\,\Omega\,(A[\nu]))\ra A^\S\,(\Gamma^\E\,\nu)\,(t^\A\,\con)\,(t^\D\,\omega^D) \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Delta^\E\,(\sigma\circ\nu) = \sigma^\S\,(\Gamma^\E\,\nu) \\
  & (t:\Tm\,\Gamma\,A)^\E && : (\nu:\Tms\,\Omega\,\Gamma)\ra A^\E\,\nu\,(t[\nu]) = t^\S\,(\Gamma^\E\,\nu)
\end{alignat*}
The definition is analogous to that of $\blank^\R$. One difference is
that for $\U$ we have an additional transport. The type that we need
is $(\alpha:a^\A\,\con)\ra (a^\D\,\omega^D\,\alpha)$. The expression
$\coe\,({a^\C\,\id}^{-1})\,\alpha$ has type
$\Tm\,\Omega\,(\El\,a)$. Now, instead of applying the standard
interpretation, we apply the logical predicate interpretation:
$(\coe\,({a^\C\,\id}^{-1})\,\alpha)^\D\,\omega^D$ has type
$a^\D\,\omega^\D\,(\alpha^\A\,\con)$ which is almost right, so we need
to coerce along $\alpha^\C\,\id$ which witnesses $\alpha^\A\,\con =
\alpha$, and we are finished. For details see Appendix A
and the Agda formalisation.

@@

\draw (m-2-2.north west) rectangle (m-17-6.south west) node[pos=.5] {\rotatebox{90}{Appendix A}};

\draw (m-2-15.north west) rectangle (m-17-18.south west) node[pos=.5] {\rotatebox{90}{Appendix A}};

@@

\draw (m-2-2.north west) rectangle (m-5-14.south west) node[pos=.5] {Appendix B};

\draw (m-9-2.north west) rectangle (m-10-14.south west) node[pos=.5] {Appendix B};
\draw (m-12-2.north west) rectangle (m-12-14.south west) node[pos=.5] {Appendix B};

@@

There are two ways to present the model: by rows or by columns of the
table. Describing the model by rows is the usual way of saying how
contexts are given in the model, then how types are given, then
substitutions, terms, the empty context, context extension and so on.
Describing the model by columns means defining operations which
interpret the syntax, and later operations can depend on previous
ones. In Sections \ref{sec:algebras}--\ref{sec:eliminator} we used the
column-based method to describe the first four columns of this
table. These were given by $\blank^\A$, $\blank^\D$, $\blank^\M$, and
$\blank^\S$ (their full definition can be found in Appendix A).

In this section we follow the row-based approach and describe the rows
of the model informally, while giving the full definition of the first
4 rows ($\Con$, $\Ty$, $\Tms$, $\Tm$) and the rows for $\U$, $\El$ and
$\Pi$ in Appendix B. These are the most interesting parts of the
construction. The rest of the construction is tedious and not very
enlightening. The first 4 columns are fully formalised in Agda and
most parts of the category-columns from the CwF-columns are also
formalised.

@@
